\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-14 article.tex
sdf

\begin{figure}[H]
  \centering
  \begin{minipage}[b]{0.69\textwidth}
    \begin{minted}[bgcolor=LightGray]{bash}
$ time dune exec build -- tests/sha256.ll -a greedy
38,72s user 0,07s system 99% cpu 38,833 total
$ time dune exec build -- tests/sha256.ll -a linear
38,72s user 0,04s system 99% cpu 38,813 total
$ time dune exec build -- tests/sha256.ll -a simple
120,55s user 0,07s system 99% cpu 2:00,74 total
    \end{minted}
    \caption{Runtime over single runs of compilation}\label{fig:build-time}
  \end{minipage}
\end{figure}

@@

To map each LLVM-{}- instruction to what amounts to the 'core' function in x86, consider Table \ref{tab:trans}. Instructions for which the size varies, the suffix is denoted with \texttt{x} instead. This is then replaced with the appropriate size suffix depending on the size denoted in the LLVM-{}- instruction. For example, most binary operations move the left-hand operand into \texttt{\%rax} and the right-hand operand into  \texttt{\%rcx}, after which the associated instruction from the table is performed. For x86, most operations are done in place, so for the most part \texttt{\%rax} is modified and moved to the destination after.

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
add      & addx \%src, \%dst & \{\}\\
alloca   & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
allocan  & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
and      & andx \%src, \%dst & \{\}\\
ashr     & sarx \%src, \%dst & \{\}\\
bitcast  & movx \%src, \%dst & \{\}\\
call     & \textrm{\textit{see \ref{impl-call}}} & \{\texttt{\%rax}, \texttt{\%rcx}, \texttt{\%rdx}, \texttt{\%rsi}, \texttt{\%r8-\%r11}\}\\
gep      & \textrm{\textit{see \ref{impl-gep}}} & \{\}\\
icmp     & cmpx, set & \{\texttt{\%rFLAGS}\}\\
load     & movx (\%src), \%dst & \{\}\\
lshr     & shrx \%src, \%dst & \{\}\\
mul      & imulx \%src, \%dst & \{\texttt{\%rax}\}\\
or       & orx \%src, \%dst & \{\}\\
phi      & \textrm{\textit{see \ref{impl-phi}}} & \{\}\\
ptrtoint & movx \%src, \%dst & \{\}\\
sdiv     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
select   & cmpx, cmoveq & \{\}\\
sext     & movx \%src, \%dst & \{\}\\
shl      & shl \%src, \%dst & \{\}\\
srem     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
store    & movx \%src, (\%dst) & \{\}\\
sub      & subx \%src, \%dst & \{\}\\
trunc    & movx \%src, \%dst & \{\}\\
udiv     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
urem     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
xor      & xorx \%src, \%dst & \{\}\\
zext     & movx \%src, \%dst & \{\}\\
 \hline
\end{tabular}
\caption{LLVM-{}- instruction to x86 translation table}\label{tab:trans}
\end{table}

@@

@@

Certainly the selection of benchmarks could also have been broader, with the cast majority just showing how Greedy is considerably slower. Not to mention more benchmarks with support for the Tiger compiler that was written the year prior. Likewise, some more detailed profiling of the entire compilation would have been interesting. Some short descriptions of time taking during certain phases are included (see Figure \ref{fig:lva-time} and \ref{fig:build-time}) although a far more detailed report could've been made with instrumentation frameworks like Landmarks \cite{landmarks} as the time spent during compile time is quite significant, prticularly for JIT.

>>>>

\begin{figure}[H]
  \centering
  \begin{minipage}[b]{0.69\textwidth}
    \begin{minted}[bgcolor=LightGray]{bash}
$ time dune exec build -- tests/sha256.ll -a greedy
38,72s user 0,07s system 99% cpu 38,833 total
$ time dune exec build -- tests/sha256.ll -a linear
38,72s user 0,04s system 99% cpu 38,813 total
$ time dune exec build -- tests/sha256.ll -a simple
120,55s user 0,07s system 99% cpu 2:00,74 total
$ time dune exec build -- tests/sha256.ll -a briggs
119,86s user 0,05s system 99% cpu 2:00,03 total
    \end{minted}
    \caption{Runtime over single runs of compilation}\label{fig:build-time}
  \end{minipage}
\end{figure}

@@

To map each LLVM-{}- instruction to what amounts to the 'core' function in x86, consider Table \ref{tab:trans}. Instructions for which the size varies, the suffix is denoted with \texttt{x} instead. This is then replaced with the appropriate size suffix depending on the type noted in the LLVM-{}- instruction. For example, most binary operations move the left-hand operand into \texttt{\%rax} and the right-hand operand into  \texttt{\%rcx}, after which the associated instruction from the table is performed. For x86, most operations are done in place, so for the most part \texttt{\%rax} is modified and moved to the destination after.

\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
add      & addx \%src, \%dst & \{\}\\
alloca   & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
allocan  & subq \$n, \%rsp & \{\texttt{\%rsp}\}\\
and      & andx \%src, \%dst & \{\}\\
ashr     & sarx \%src, \%dst & \{\}\\
bitcast  & movx \%src, \%dst & \{\}\\
call     & \textrm{\textit{see \ref{impl-call}}} & \{\texttt{\%rax}, \texttt{\%rcx}, \texttt{\%rdx}, \texttt{\%rsi}, \texttt{\%r8-\%r11}\}\\
gep      & \textrm{\textit{see \ref{impl-gep}}} & \{\}\\
icmp     & cmpx, set & \{\texttt{\%rFLAGS}\}\\
load     & movx (\%src), \%dst & \{\}\\
lshr     & shrx \%src, \%dst & \{\}\\
mul      & imulx \%src, \%dst & \{\texttt{\%rax}\}\\
or       & orx \%src, \%dst & \{\}\\
phi      & \textrm{\textit{see \ref{impl-phi}}} & \{\}\\
ptrtoint & movx \%src, \%dst & \{\}\\
sdiv     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
select   & cmpx, cmoveq & \{\texttt{\%rFLAGS}\}\\
sext     & movx \%src, \%dst & \{\}\\
shl      & shl \%src, \%dst & \{\}\\
srem     & idivx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
store    & movx \%src, (\%dst) & \{\}\\
sub      & subx \%src, \%dst & \{\}\\
trunc    & movx \%src, \%dst & \{\}\\
udiv     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
urem     & divx \%src, \%dst & \{\texttt{\%rax}, \texttt{\%rdx}\}\\
xor      & xorx \%src, \%dst & \{\}\\
zext     & movx \%src, \%dst & \{\}\\
 \hline
\end{tabular}
\caption{LLVM-{}- instruction to x86 translation table}\label{tab:trans}
\end{table}

@@

\subsubsection{Translating terminators} \label{impl-term}

As for implementing terminators, the simplest is \texttt{unreachable}, as this literally doesn't evaluate to anything.  Next is \texttt{ret}, which is more involved, in that it needs to concern itself with calling conventions.
Although these are fairly straight forward, expecially since functions generally apply the same operations in the prologue by pushing all callee-saved registers to the stack and then, occasionally, subctracting the amount of spilled variables from \texttt{\%rsp}.
The function epilogue doesn't have to concern itself with the number of spiller variables though, as the stack pointer is saved to \texttt{\%rbp} and simply restored from that.
The return value (if specified) is moved to the return register which is \texttt{\%rax} according to the System V AMD64 ABI \cite[26]{sysv}.

@@

Certainly the selection of benchmarks could also have been broader, with the cast majority just showing how Greedy is considerably slower. Not to mention more benchmarks with support for the Tiger compiler that was written the year prior. Likewise, some more detailed profiling of the entire compilation would have been interesting. Some short descriptions of time taking during certain phases are included (see Figure \ref{fig:lva-time} and \ref{fig:build-time}) although a far more detailed report could've been made with instrumentation frameworks like Landmarks \cite{landmarks} as the time spent during compile time is quite significant, prticularly for runtime environments that depend on JIT, but also for AOT compilers as long compile times can negatively affect productivity.
